\special{dvipdfmx:config z 0}
\documentclass{article}
\usepackage{marcythm}

\title{Undergraduate Complexity Theory \\ Lecture 26: Beyond Worst-Case Analysis}
\author{Marcythm}
% \date{\today}
\date{July 29, 2022}

\begin{document}
\maketitle{}

\section{Lecture Notes}

{\bf Assume \(\P \neq \NP\) for the rest lectures.}

So \(\prob{3SAT} \notin \P\). Possibilities:
\begin{enumerate}
  \item Allow more than \(\poly(n)\) time to solve all instances.
  \item Relax correctness: ``Approx Algos'', maybe \(\exists\) poly-time \(A\) s.t. \(\forall \phi \in \prob{3SAT}\), \(A(\phi)\) outputs an assignment satisfying 90\% of clauses.
  \item Look for poly-time algo that correct on ``most'' inputs, may be \(\geq 99\%\) of all 3CNFs.
\end{enumerate}
Dream: Show hardness for 1, 2, 3, just assuming \(\P \neq \NP\). \\
Backup Dream: Make one new basic assumption, try to derive many consequences.

\begin{mdframed}
  For point 1: \\
  E.T.H.: \(\exists \delta_0 > 0: \prob{3SAT} \notin \TIME(2^{\delta_0 n})\), active field of research right now, conseqs of this in next lec. \\
  If ETH is true, then no algo solve \prob{LCS} in \(O(n^2)\) time. \\
  \(\prob{\~ETH} \implies \BPP = \P\) (mentioned in lec 22).
\end{mdframed}

\begin{mdframed}
  For point 2: partially realized. \\
  PCP Theorem ('93): \(\P \neq \NP \implies \) not exists poly-time \(A\) s.t. \(\forall \phi \in \prob{3SAT}\), \(A(\phi)\) outputs assignment satisfying \(\geq 99.999\ldots\%\) of clauses. \\
  Håstad'99 get this fraction down to \(7/8 + \eps\). (optimal hardness result, since actually exists an algo that always output assignment satisfying \(7/8\) of clauses.)
\end{mdframed}

\begin{fact}
  \(\exists\) efficient randomize algo \(A\) s.t. \(\E\bracks{\text{\# clauses satisfied by \(A(\phi)\)}} \geq 7m/8\).
\end{fact}

\begin{lemma}
Let \(\phi(x)\) be a E3CNF, let \(a \in \setof{0, 1}^n\) be a uniformly random assignment, then
  \[ \E\bracks{\text{\# clauses sat'd by \(a\)}} = \frac{7}{8} m \]
\end{lemma}

\begin{proof}
  Let \(I_j = \begin{cases}
    1, & \text{if \(a\) sats the \(j\)th clause of \(\phi\)} \\
    0, & \text{otherwise}
  \end{cases}\), \(j = 1 \ldots, m\). Then
  \[ \E[\#] = \E[I_1] + \E[I_2] + \cdots + \E[I_n] = \sum_{i=1}^m \Pr[\text{\(i\)th clause is sat'd by \(x = a\)}] \]
  For each clause, it's in the form \(x_i \vee x_j \vee \ol{x_k}\), so the probability it's sat'd is \(7/8\). Thus
  \[ \E[\#] = \frac{7}{8} m \]
\end{proof}

Min-Bisection: Given \(G\), find \(S \subseteq V, |S| = n/2\) s.t. \# edges between \(S\) and \(\ol{S}\) is minimized. \NP-hard in '70s, and in 2017: no known poly-time algo that achieves \(\leq C \times \min\) (but can achieve \(\leq \log^2 n \times \min\)). Assume \(\P \neq \NP\), we can't rule outpoly-time algo getting factor \(1 + \eps\).

\begin{remark}
  The reductions between (\NP-complete) problems only preserve exact satisfiability, but not preserve approximation quality.
\end{remark}

Worst-Case Hardness: \(\forall\) poly-time \prob{SAT} solvers, \(\exists\) poly-time formula generator s.t. {\tt Solve(Gen(n))} fails (for sufficiently large \(n\)).

Average-Case Hardness: \(\exists\) poly-time formula generator, \(\forall\) poly-time \prob{SAT} solvers {\tt Solve(Gen(n))} fails (with high probability) (for sufficiently large \(n\)). Here the generation algo always allowed to be randomized, since otherwise the solver can hard-code the solution.

Average-Case Hardness is always desirable, for cryptography.

\begin{definition}[Uniformly Random \prob{E3SAT} Instances]
  \(n = \text{\# vars}\), \(\Delta = \text{``clause density''}\), \(m := \Delta n\).
  \[ \phi = \text{Choose \(m\) random clauses independently, each clause is uniformly chosen from all \(2^3 \binom{n}{3}\) possibilities} \]
\end{definition}

Q0: Is \(\phi\) likely to be sat or unsat? Depends on \(\Delta\).

\begin{exercise}
  If \(\Delta > 5.2\), then \(\phi\) is exponentially unlikely to be sat.
\end{exercise}

\begin{theorem}
  If \(\Delta \leq 0.16\), then \(\phi\) is exponentially unlikely to be unsat.
\end{theorem}

\begin{conjecture}
  \(\exists\) ``phase transition'' at \(\Delta \approx 4.2667\), i.e. a sharp decrease of probability.
\end{conjecture}

\begin{exercise}[Chernoff bound]
  If \(\Delta \geq 10\)ish, w.h.p. \(\phi\) is unsat, and every assignment \(a \in \setof{0, 1}^n\)  sats \(\leq 7/8 + \eps\) fraction of constraints.
\end{exercise}

Algo hardness: if \(\Delta < 4.2667\cdots\): solver should find a sat assignment.
2017: poly-time algo that provably works when \(\Delta \leq 3.52\).
Algos that seem to work in poly-time for \(\Delta \leq 4.266\cdots\). \\
If \(\Delta > 4.2667\cdots\): solver should give a proof that \(\phi\) is unsat.

\begin{definition}
  A \(\Delta\prob{-3SAT}\)-Refuter is a poly-time algo that:
  given any \(\phi\), either outputs ``unsat'' or ``no comment'';
  is never wrong.
  Given random \(\phi\) with param \(\Delta\), \(\Pr\bracks{\text{\(A(\phi)\) = unsat}} \geq 99\%\).
\end{definition}

In 2017: do \(\Delta-\prob{-3SAT}\)-Refuters exist for \(\Delta = C\)? Unknown. But for \(\Delta = \sqrt{n}\), it exists.

\begin{hypothesis}[Feige's Hypothesis]
  \(\forall\) constant \(\Delta\), \(\Delta-\prob{3SAT}\)-Refuter don't exist.
\end{hypothesis}

This hypothesis directly implies that \(\P \neq \NP\) (or there will exist poly-time algo for \prob{3SAT}, thus such refuter exists), Håstad's result (left as exercise), and no poly-time approx for Min-Bisection problem achieving \(\leq 4/3\).

\end{document}
